Demes: a standard format for demographic models

### 2022-08-02 paper.tex
Move content into main text and rejig

Below we provide a brief overview of the population genetics models
that Demes supports and the components of the Demes infrastructure.
We then highlight an example usage, including visualization and simulation
of a multi-population demographic model.
More detailed descriptions of the population genetics model, parsers, scope, and
additional examples are given in the Appendix.
Complete technical details of the MDM and HDM, and the responsibilities
of the parser are provided in the online Demes specification
(\url{https://popsim-consortium.github.io/demes-spec-docs/}).

@@

The Demes Human Data Model (HDM) is focused on efficient human
understanding and avoiding errors.
We have adopted the widely used YAML format~\citep{ben2009yaml} as the
primary interface for writing and interchanging demographic models
(see Appendix~\ref{sec:appendix-spec} for details about our choice of
YAML).
Demographic models written in YAML format provide information about global
features of the model (such as time units and generation times),
populations (as ``demes'') and their existence intervals (as ``epochs''),
and gene flow between populations
(as continuous ``migrations'' or instantaneous ``pulse'' events)
(e.g., Figures~\ref{fig:IM}~and~\ref{fig:showcase}).
The HDM encourages human understanding by avoiding
redundancy in the description where possible and by providing a mechanism for
specifying default values that are inherited hierarchically.

@@

%\subsection*{Machine Data Model}
A Demes model in MDM form is also in valid HDM form, so parsing an already
resolved model does not change the model. Furthermore, this allows users to be
more verbose and write values even when they are not required by the HDM.
Being explicit may be preferred for reasons of clarity, or if the user is
unsure about how an HDM model will be resolved into an MDM model.
Although we note that in the latter case, users can simply look at the
parser output.

\subsection*{The scope of Demes}

The Demes specification is static by design---we wish to unambiguously
describe a demographic model with a concrete set of parameters.
This simplicity means that we cannot directly specify parameter distributions
or estimated confidence intervals for those parameters.
While it is not difficult to imagine extending the specification in ways that
would allow this (Appendix~\ref{sec:appendix-scope}), it is not immediately
clear that the benefits are worth the increased parser complexity.
It is important to note, however, that Demes may be used in applications that
include additional population genetic processes outside of what is explicitly
modeled in the specification, such as interpreting population sizes as
carrying capacities, implementations of hard selection, or layering more
complicated mating or spatial structure. The Demes specification therefore
provides a basic model that can be elaborated on where necessary.

\section*{Example: an isolation with migration model}

@@

\section*{Discussion}
Stable and healthy software ecosystems require standard interchange
formats, allowing for the development of high-quality and long-lasting
tools that produce and consume the standard.
Demographic models are a key part of population genetics research,
and to date the transfer of inferred models to downstream simulations
has been \textit{ad-hoc}, and conversions between the many different ways
of expressing such models is both labor intensive and fraught with errors.
The proposed Demes standard is an attempt to bridge this gap
between inference and simulation, and also to provide the foundations
for a sustainable ecosystem of tools built around this data model.
Table~\ref{tab:software} shows some initial infrastructure that we have
built as part of developing Demes, but many other useful tools
can be envisaged that consume, transform, or produce this format.

@@

\section*{Acknowledgments}
Graham Gower was supported by a Villum Fonden Young Investigator award to Fernando Racimo (project no. 00025300).
Ryan Gutenkunst was supported by the National Institute of General Medical Sciences of the National Institutes of Health (R01GM127348 to RNG).
Matthew Hartfield is supported by a NERC Independent Research Fellowship (NE/R015686/1).

@@

The Demes specification is a formal data model for describing
the properties of populations over time,
along with some metadata and provenance information.
The data model is based on the ubiquitous JSON~\citep{bray2017javascript}
standard, and formally defined using
JSON Schema~\citep{wright2020json}.
Along with the schema, full technical details of the
of the model are provided in the
online specification document
(\url{https://popsim-consortium.github.io/demes-spec-docs/}).
This specification rigorously defines the data model,
fully describing the entities and their
relationships, and the required behavior of implementations.
Since the online specification is definitive,
we will not recapitulate the details
here, but instead focus on the high level properties of the model and
the rationale behind key design decisions.

Below we detail the population genetics model in
Appendix~\ref{sec:appendix-pop-gen-model},
the high-level, human-readable model specification in
Appendix~\ref{sec:appendix-spec}, demographic model parsers and their
intended behavior in Appendix~\ref{sec:appendix-parsers}, and the
intended scope of Demes in Appendix~\ref{sec:appendix-scope}.
The extended figures in Appendix~\ref{sec:appendix-figures} provide
additional examples and illustrations of Demes usage.

@@

\section{Human Data Model}\label{sec:appendix-spec}

The Demes Human Data Model (HDM) is focused on the efficiency of
human understanding and the avoidance of errors. We have adopted the widely
used YAML format~\citep{ben2009yaml} as the recommended means of interchanging
Demes models (e.g., Figures~\ref{fig:IM}~and~\ref{fig:tennessen}). YAML is a data
serialization language with an emphasis on simplicity and which interoperates well
with JSON (indeed, YAML 1.2 is a super set of JSON). We chose YAML over JSON
because although JSON is an excellent format for data interchange, it is
ill-suited for human understanding or manipulation. We also considered other
declarative data exchange formats such as TOML,
but chose YAML because of equivalence with JSON,
its popularity and good software support.
Since the Demes data model is defined in JSON Schema,
however, there is no formal dependency on YAML and implementations may choose
to use JSON directly if they wish (e.g., for greater efficiency).

Structurally, the HDM encourages human understanding
by avoiding redundancy in the description where possible and by providing a
mechanism for specifying default values that are inherited hierarchically.
Figure~\ref{fig:IM} shows an example Demes model expressed in both the
HDM and Machine Data Model (MDM) forms (both in terms of YAML syntax).
For values that repeat
across fields, defaults may be used to implicitly assign default values to
fields of the given type.  A default is superseded by an explicitly provided
value if given. Size values are inherited naturally following the
progression of time. For example, if an epoch \texttt{start\_size} is not
provided (neither directly, nor via a defaults section),
it is assumed to be equal to the \texttt{end\_size} of the previous
epoch. This also means that the first epoch of each population must specify the
initial size (or it must be provided in a defaults section).

\section{Scope of the specification}
\label{sec:appendix-scope}

\subsection{Static models, not parameterized models}
\label{sec:appendix-static}

@@

\subsection{Population-level features, not genome features}
\label{sec:appendix-features}

Demes is designed to interface with a large number of simulation and inference
methods, so we have restricted the specification to only describe
demographic features at the population level.
Features of the underlying genome biology are omitted,
including mutation and recombination rates,
genome annotations, ploidy, and so on.
Selection and dominance models are absent, as discussed in
Sec.~\ref{sec:appendix-pop-gen-model}, and we do not
specify how individuals are sampled.

The model is, by design, very limited. We do not specify details of
individuals or of the processes that happen within populations,
because such details are necessarily complicated, vary greatly by
application, and attempts to encompass such a broad range of biological
processes are unlikely to succeed.
The narrow focus on individuals and their groupings into populations
should ensure that the Demes standard is stable, and not subject to
continual development as more and more elaborations of basic processes
are added. Demes is intended to provide \emph{part} of a
simulation model specification; other parts of the specification,
such as sampling strategies, or parameters that vary by time
or population can \emph{refer to} this well defined and unambiguous
description of the demography.

>>>>

Here we provide a brief overview of the population genetics models
that Demes supports and the components of the Demes infrastructure.
Complete technical details of the MDM and HDM, and the responsibilities
of the parser are provided in the online Demes specification
(\url{https://popsim-consortium.github.io/demes-spec-docs/}).
This specification rigorously defines the data model,
fully describing the entities and their
relationships, and the required behavior of implementations.
Since the online specification is definitive,
we will not recapitulate the details
here, but instead focus on the high level properties of the model and
the rationale behind key design decisions.

@@

The Demes Human Data Model (HDM) is focused on efficient human
understanding and avoiding errors.
We have adopted the widely used YAML format~\citep{ben2009yaml} as the
primary interface for writing and interchanging demographic models
(see Appendix~\ref{sec:appendix-spec} for rationale).
Demographic models provide information about global
features of the model (such as time units and generation times),
populations (as ``demes'') and their existence intervals (as ``epochs''),
and gene flow between populations
(as continuous ``migrations'' or instantaneous ``pulse'' events).
Fig~\ref{fig:IM} shows an isolation-with-migration in HDM format.

Structurally, the HDM encourages human understanding
by avoiding redundancy in the description where possible and by providing a
mechanism for specifying default values that are inherited hierarchically.
For values that repeat
across fields, defaults may be used to implicitly assign default values to
fields of the given type.  A default is superseded by an explicitly provided
value if given. Size values are inherited naturally following the
progression of time. For example, if an epoch \texttt{start\_size} is not
provided (neither directly, nor via a defaults section),
it is assumed to be equal to the \texttt{end\_size} of the previous
epoch. This also means that the first epoch of each population must specify the
initial size (or it must be provided in a defaults section).

Avoiding redundancy in this way reduces the cognitive load on readers,
by highlighting necessary parameters which may be otherwise be obscured.
It is not necessary---or indeed recommended---that all models are expressed
in a maximally concise form, and we wholeheartedly endorse
the explicit statement of parameters where it increases model legibility.

@@

\subsection*{Scope of the specification}

A primary design goal of Demes is to provide a means of
unambiguously communicating the results of demographic
model inferences to population genetic simulators.
Since demography is defined in terms of groups of individuals
and these groupings are influenced by genetics,
it is difficult to find a simple definition that separates
the two. Thus, we have attempted to be pragmatic,
limiting the features that we include in Demes to
those that are (or can reasonably be expected to)
in practise regarded as part of a demographic model.

Thus, the model is limited to features that we can expect many different
demographic inference and simulation methods to share. The specification only
describes demographic features at the population level. Features of genome
biology are out of scope, including mutation and recombination rates,
genome annotations, ploidy, and so on. Selection and dominance models are
absent, as discussed in Appendix.~\ref{sec:appendix-pop-gen-model}.
It is important to note, however, that Demes may be used in applications that
include additional population genetic processes outside of what is explicitly
modelled in the specification, such as interpreting population sizes as
carrying capacities, implementations of hard selection, or layering more
complicated mating or spatial structure. The Demes specification
is intended to provide a basic model that can be elaborated on where necessary.

Demes is not a standard population genetic simulation specification,
although it could be \emph{part} of one.
Since the standard is based on JSON, and JSON documents can be
arbitrarily nested, it is possible to imagine a simple specification
of genome features such as mutation and recombination rates
in which the demography is defined by an embedded Demes specification.
Features of the simulation specification (such as defining
the time and location of samples) can then \emph{refer to} the Demes
model. This design, in which we embed the Demographic model
\emph{within} a larger specification rather than adding arbitrary
and unrelated complexities \emph{to} the demography
is an essential simplification and separation of duties.

The Demes specification is static by design---we wish to unambiguously
describe a demographic model with a concrete set of parameters.
This simplicity means that we cannot directly specify parameter distributions
or estimated confidence intervals for those parameters.
While it is not difficult to imagine extending the specification in ways that
would allow this, it is not
clear that the benefits are worth the increased parser complexity
(see Appendix~\ref{sec:appendix-static}).

\section*{Example: an isolation-with-migration model}

@@

\section*{Discussion}
Stable and healthy software ecosystems require standard interchange
formats, allowing for the development of high-quality and long-lasting
tools that produce and consume the standard.
Demographic models are a key part of population genetics research,
and to date the transfer of inferred models to downstream simulations
has been \textit{ad-hoc}, and conversions between the many different ways
of expressing such models is both labor intensive and fraught with errors.
The proposed Demes standard is an attempt to bridge this gap
between inference and simulation, and also to provide the foundations
for a sustainable ecosystem of tools built around this data model.
Table~\ref{tab:software} shows some initial infrastructure that we have
built as part of developing Demes, but many other useful tools
can be envisaged that produce, consume, or transform this format.

@@

\section*{Acknowledgments}
Graham Gower was supported by a Villum Fonden Young Investigator award to Fernando Racimo (project no. 00025300).
Ryan Gutenkunst was supported by the National Institute of General Medical Sciences of the National Institutes of Health (R01GM127348 to RNG).
Matthew Hartfield is supported by a NERC Independent Research Fellowship (NE/R015686/1).
Jerome Kelleher is supported by the Robertson Foundation.

@@

The Demes specification is a formal data model for describing
the properties of populations over time,
along with some metadata and provenance information.
The data model is based on the ubiquitous JSON~\citep{bray2017javascript}
standard, and formally defined using
JSON Schema~\citep{wright2020json}.
Along with the schema, full technical details of the
of the model are provided in the
online specification document
(\url{https://popsim-consortium.github.io/demes-spec-docs/}).

@@

\section{Rationale for YAML}\label{sec:appendix-spec}

We have adopted the widely
used YAML format~\citep{ben2009yaml} as the recommended means of interchanging
Demes models (e.g., Figures~\ref{fig:IM}~and~\ref{fig:tennessen}). YAML is a data
serialization language with an emphasis on simplicity and which interoperates well
with JSON (indeed, YAML 1.2 is a super set of JSON). We chose YAML over JSON
because although JSON is an excellent format for data interchange, it is
ill-suited for human understanding or manipulation. We also considered other
declarative data exchange formats such as TOML,
but chose YAML because of equivalence with JSON,
its popularity and good software support.
Since the Demes data model is defined in JSON Schema,
however, there is no formal dependency on YAML and implementations may choose
to use JSON directly if they wish (e.g., for greater efficiency).

\section{Rationale for static models}
\label{sec:appendix-static}

@@
